---
title: "Racial Bias in Gifted and Talented Programs"
author: "Alexandra Chang"
subtitle: Data Science I (STAT 301-1)
output:
  pdf_document:
    toc: yes
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    highlight: tango
---

## Overview

The goal of this lab is to better understand the basic data structure called a "tibble," which is used throughout the `tidyverse` instead of R's traditional `data.frame`. Tibbles are data frames, but they tweak some of R's older behaviors to make life a little easier by avoiding unintentional mistakes/errors. The `tibble` package provides the framework for these opinionated data frames which make working with data in the tidyverse possible -- [`tibble` package home page](http://tibble.tidyverse.org/index.html).

## Datasets 

This lab utilizes the `mtcars` and `flights` datasets contained in the packages `datasets` (automatically loaded with base R) and `nycflights13`, respectively. For one of the problems, you may opt to use one of our other familiar datasets, `diamonds` or `mpg`, which are contained in `ggplot2`.   

## Exercises

Please complete the following exercises. Be sure your solutions are clearly indicated and that the document is neatly formatted. 

<br>

#### Load Packages

```{r, message = FALSE}
# Loading package(s)
library(tidyverse)
library(ggplot2)
```

<br>

# rcode

```{r}
## Load packages ----------------------------------------------------------
library(tidyverse)
library(ggplot2)
library(readr)

## Data import -------------------------------------------------------------
enroll_raw <- read.csv(file = "data/unprocessed/Enrollment.csv")
gat_raw <- read.csv(file = "data/unprocessed/Gifted and Talented.csv")

```

```{r}
## Prep for join -----------------------------------------------------------
intersect(names(enroll_raw),
          names(gat_raw))

enroll_raw %>% 
  count(COMBOKEY) %>% 
  filter(n > 1) # Confirm key

gat_raw %>% 
  count(COMBOKEY) %>% 
  filter(n > 1) # Confirm key


## enroll_simple -------------------------------------------------------------------------
enroll_simple <- enroll_raw %>% 
  as.tibble() %>% 
  mutate(enroll_raw, 
         state = LEA_STATE, 
         hisp_m = SCH_ENR_HI_M, hisp_f = SCH_ENR_HI_F, 
         asian_m = SCH_ENR_AS_M, asian_f = SCH_ENR_AS_F, 
         black_m = SCH_ENR_BL_M, black_f = SCH_ENR_BL_F, 
         white_m = SCH_ENR_WH_M, white_f = SCH_ENR_WH_F, 
         other_m = SCH_ENR_AM_M + SCH_ENR_HP_M + SCH_ENR_TR_M, 
         other_f = SCH_ENR_AM_F + SCH_ENR_HP_F + SCH_ENR_TR_F, 
         male = hisp_m + asian_m + black_m + white_m + other_m, 
         female = hisp_f + asian_f + black_f + white_f + other_f, 
         total = male + female) %>% # rename and collapse
  select(COMBOKEY, state, hisp_m, hisp_f, 
         asian_m, asian_f, black_m, black_f, 
         white_m, white_f, other_m, other_f, 
         male, female, total) # select

sapply(enroll_simple, class) # vector types

x <- c(3:15) # change columns 3 - 15 to numeric
enroll_simple[,x]

enroll_simple[,x] <- apply(enroll_simple[,x], 2, 
                           function(x) {as.numeric(as.integer(x))})

sapply(enroll_simple, class) # vector types after transformation

```

```{r}
## gat_filtered --------------------------------------------------------------
gat_simple <- gat_raw %>% 
  as.tibble() %>% 
  mutate(gat_raw, 
         state = LEA_STATE, 
         gat_hisp_m = SCH_GTENR_HI_M, gat_hisp_f = SCH_GTENR_HI_F, 
         gat_asian_m = SCH_GTENR_AS_M, gat_asian_f = SCH_GTENR_AS_F,
         gat_black_m = SCH_GTENR_BL_M, gat_black_f = SCH_GTENR_BL_F, 
         gat_white_m = SCH_GTENR_WH_M, gat_white_f = SCH_GTENR_WH_F,
         gat_other_m = SCH_GTENR_AM_M + SCH_GTENR_HP_M + SCH_GTENR_TR_M, 
         gat_other_f = SCH_GTENR_AM_F + SCH_GTENR_HP_F + SCH_GTENR_TR_F, 
         gat_male = gat_hisp_m + gat_asian_m + gat_black_m + gat_white_m + gat_other_m, 
         gat_female = gat_hisp_f + gat_asian_f + gat_black_f + gat_white_f + gat_other_f, 
         gat_total = gat_male + gat_female) %>% # rename and collapse
  select(COMBOKEY, state, SCH_GT_IND, 
         gat_hisp_m, gat_hisp_f, 
         gat_asian_m, gat_asian_f, 
         gat_black_m, gat_black_f, 
         gat_white_m, gat_white_f, 
         gat_other_m, gat_other_f, 
         gat_male, gat_female, gat_total) # select

sapply(gat_simple, class) # vector types

y <- c(4:16) # change columns 4 - 16 to numeric
gat_simple[,y]

gat_simple[,y] <- apply(gat_simple[,y], 2, 
                        function(x) {as.numeric(as.integer(x))}) # change columns 3 - 15 to numeric

sapply(gat_simple, class) # vector types after transformation

```

```{r}
## inner_join() ------------------------------------------------------------
join1 <- enroll_simple %>% 
  inner_join(gat_simple, 
             by = "COMBOKEY", 
             suffix = c("", "")) # inner_join()

```

```{r}
## enroll_gat --------------------------------------------------------------
enroll_gat <- join1 %>% 
  filter(SCH_GT_IND == "Yes", 
         gat_total > 0) %>% # filter schools with GAT program
  filter(male > 0, 
         female > 0) %>% # filter schools with males AND females
  mutate(prop_gat = gat_total / total) %>% 
  filter(prop_gat < 1) %>% 
  select(COMBOKEY, state, 
         gat_asian_m, asian_m, gat_asian_f,asian_f,
         gat_black_m, black_m, gat_black_f, black_f, 
         gat_hisp_m, hisp_m, gat_hisp_f, hisp_f, 
         gat_white_m, white_m, gat_white_f, white_f, 
         gat_other_m, other_m, gat_other_f, other_f, 
         gat_male, male, gat_female, female, 
         gat_total, total) # select

sapply(enroll_gat, class) # vector types

```

```{r}
## Race-Ethnicity (RE) population by school -------------------------------------------
RE_pop <- enroll_gat %>% 
  as_tibble() %>% 
  mutate(enroll_gat, 
         asian = asian_m + asian_f,
         black = black_m + black_f,
         hisp = hisp_m + hisp_f,
         white = white_m + white_f,
         other = other_m + other_f, # enrollment totals by race
         bipoc = asian + black + hisp + other) %>% 
            # BIPOC vs white
  mutate(enroll_gat, 
         gat_asian = gat_asian_m + gat_asian_f,
         gat_black = gat_black_m + gat_black_f,
         gat_hisp = gat_hisp_m + gat_hisp_f,
         gat_white = gat_white_m + gat_white_f,
         gat_other = gat_other_m + gat_other_f, # GAT totals by race
         gat_bipoc = gat_asian + gat_black + gat_hisp + gat_other) %>% 
            # GAT BIPOC vs GAT white
  mutate(enroll_gat, 
         prop_gat_asian = gat_asian / asian, 
         prop_gat_black = gat_black / black, 
         prop_gat_hisp = gat_hisp / hisp, 
         prop_gat_white = gat_white / white, 
         prop_gat_other = gat_other / other, 
         prop_gat_bipoc = gat_bipoc / bipoc) %>% # GAT:population ratios
  mutate(enroll_gat, 
         prop_gat_total = gat_total / total) %>% # proportion of students in GAT programs
  na.omit() %>% # omit NaN
  select(state, prop_gat_asian, gat_asian, asian, 
         prop_gat_black, gat_black, black, 
         prop_gat_hisp, gat_hisp, hisp, 
         prop_gat_white, gat_white, white, 
         prop_gat_other, gat_other, other, 
         prop_gat_bipoc, gat_bipoc, bipoc,
         prop_gat_total, gat_total, total)

# I am filtering out schools were more than 50% of students are in
# GAT programs. At some schools, ~100% of students are in GAT programs, 
# leading me to question, are these truly "gifted and talented" programs 
# given their lack of selectivity and exclusivity?
RE_pop <- RE_pop %>% 
  mutate(prop_gat_asian =
           round(RE_pop$prop_gat_asian, digits = 2), 
         prop_gat_black =
           round(RE_pop$prop_gat_black, digits = 2), 
         prop_gat_hisp =
           round(RE_pop$prop_gat_hisp, digits = 2), 
         prop_gat_white =
           round(RE_pop$prop_gat_white, digits = 2), 
         prop_gat_other =
           round(RE_pop$prop_gat_other, digits = 2), 
         prop_gat_bipoc =
           round(RE_pop$prop_gat_bipoc, digits = 2), 
         prop_gat_total =
           round(RE_pop$prop_gat_total, digits = 2)) %>% # rounding to make my life easier
  filter(prop_gat_total <= 0.5) %>%  # filter out if > 50%
  filter(prop_gat_asian <= 1, 
         prop_gat_black <= 1, 
         prop_gat_hisp <= 1, 
         prop_gat_white <= 1, 
         prop_gat_other <= 1, 
         prop_gat_bipoc <= 1) # filter out if > 100%

# For my visualizations, I want to compare the proportion of students of color in GAT
# programs to the proportion of white students. prop_gat_white will be the independent variable.
RE_pop <- RE_pop %>% 
  arrange(state, prop_gat_white, gat_white, white,
          prop_gat_bipoc, gat_bipoc, bipoc,
          prop_gat_asian, gat_asian, asian, 
          prop_gat_black, gat_black, black, 
          prop_gat_hisp, gat_hisp, hisp, 
          prop_gat_other, gat_other, other, 
          prop_gat_total, gat_total, total)
```




#### Visualizations

```{r}
ggplot(RE_pop, aes(prop_gat_total, prop_gat_asian)) + 
  geom_jitter(color = "grey25", size = 0.5) + 
  geom_smooth(color = "gold") +
  geom_abline(intercept = 0, slope = 1, size = 1) # asian vs total
```

```{r}
ggplot(RE_pop, aes(prop_gat_total, prop_gat_black)) + 
  geom_jitter(color = "grey25", size = 0.5) + 
  geom_smooth(color = "deeppink") + 
  geom_abline(intercept = 0, slope = 1, size = 1) # black vs total
```

```{r}
ggplot(RE_pop, aes(prop_gat_total, prop_gat_hisp)) + 
  geom_jitter(color = "grey25", size = 0.5) + 
  geom_smooth(color = "darkolivegreen1") + 
  geom_abline(intercept = 0, slope = 1, size = 1) # hisp vs total
```

```{r}
ggplot(RE_pop, aes(prop_gat_total, prop_gat_white)) + 
  geom_jitter(color = "grey25", size = 0.5) + 
  geom_smooth(color = "red") + 
  geom_abline(intercept = 0, slope = 1, size = 1) # white vs total
```

```{r}
ggplot(RE_pop, aes(prop_gat_total, prop_gat_other)) + 
  geom_jitter(color = "grey25", size = 0.5) + 
  geom_smooth(color = "cornflowerblue") + 
  geom_abline(intercept = 0, slope = 1, size = 1) # other race vs total
```

```{r}
ggplot(RE_pop, aes(prop_gat_total, prop_gat_bipoc)) + 
  geom_jitter(color = "grey25", size = 0.5) + 
  geom_smooth(color = "blueviolet") + 
  geom_abline(intercept = 0, slope = 1, size = 1) # BIPOC vs total

```


```{r}
RE_pop %>% 
  ggplot() + 
  geom_smooth(aes(prop_gat_total, prop_gat_asian), color = "gold", se = F) + 
  geom_smooth(aes(prop_gat_total, prop_gat_black), color = "deeppink", se = F) +
  geom_smooth(aes(prop_gat_total, prop_gat_hisp), color = "darkolivegreen1", se = F) + 
  geom_smooth(aes(prop_gat_total, prop_gat_white), color = "red", se = F) + 
  geom_smooth(aes(prop_gat_total, prop_gat_other), color = "cornflowerblue", se = F) + 
  geom_smooth(aes(prop_gat_total, prop_gat_bipoc), color = "blueviolet", se = F) + 
  geom_abline(intercept = 0, slope = 1, size = 1) +
  xlab("Proportion of students in GAT") + 
  ylab("Proportion of students in GAT by race")
```

